The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-31 manuscript/5pillars.Rmd
spellcheck american english

* Correspondance: m.ziemann@deakin.edu.au

| Author | ORCID |
| --- | --- |
| Mark Ziemann | 0000-0002-7688-6974 |
| Pierre Poulain | 0000-0003-4177-3619 |
| Anusuiya Bora | 0009-0006-2908-1352 |

@@

In research, computational reproducibility is the ability to use the materials from a past study
(such as data, code and documentation) to regenerate the outputs including figures and tables to
confirm the study's findings @Goodman2016-wi.
Reproducibility is only the first step towards overall reliability, for example a study may be
reproducible, but suffer from analytical problems that invalidate the results.
If a study is reproducible, then at least these analytical issues can be more easily identified and
rectified.
Therefore, reproducibility is a necessary stepping-stone in the framework of cumulative science.
Reproducibility should enhance the overall reliability of computational research
including replicability and robustness.
Replicability being defined as the ability to repeat the entire study and come up with similar overall
results.
Robustness refers to the ability for the findings to be generalisable to other situations
@Nosek2022-zc.

@@

For example, in 2006 an article entitled "Genomic signatures to guide the use of chemotherapeutics"
generated a great deal of interest as it was an early application of high throughput transcriptomics
in the prediction of individual patient responses to different chemotherapies @Potti2006-yk.
After observing some unusual features of the patient group, Baggerly and Coombes attempted
reproduction of some of the key findings @Baggerly2010-nr.
Without the help of scripted workflows to guide the re-analysis, the team used forensic bioinformatic
techniques to piece together how the dataset was originally analysed.
Their investigations found a litany of issues.
Firstly, the labeling of patients in the test set as "responders" or "non-responders" had been reversed
in the original analysis.
Secondly, some of the patients were included more than once (some up to four times) in the analysis,
likely to cause major distortions in results.
Confusingly, some of the reused data has inconsistent grouping; ie: some of the samples are labeled
both sensitive and resistant.
Additional errors include two cases where results (charts) were ascribed to the wrong drug.
Baggerly and Coombes highlight that such mistakes can inadvertently occur when conducting unscripted
data analysis such as using spreadsheets, and these problems can be obscured by a lack of documentation.
The article underwent two corrigenda, but was untimately retracted in 2011, as the authors were not
able to reproduce the findings themselves due to "corruption of several validation data sets"
@Potti2011-ut.
As the array findings were the basis for clinical trials where patients were allocated to treatments,
the flawed data analysis may have harmed patients given the wrong drug in the
period 2007-2010.
In 2010, Duke University terminated the trials and suspended the lead author, Dr Anil Potti, who
later resigned.
Duke was served eight lawsuits by families of affected patients seeking compensation for exposure to
harmful and unnecessary chemotherapy, which were settled out of court @Dyer2015-my.
This worst-case scenario emphasizes that computational reproducibility is crucial for
translating bioinformatics research into real-world outcomes.

@@

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
[@Kurtzer2017-zq], Docker remains the most widely used containerisation system (according to GitHub
stars as of August 2023).
Docker has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei; @Mangul2019-xp; @Krafczyk2021-xv].

@@

Literate programming also provides an opportunity to record sanity checks in an output report.
In R, commands like `dim()`, `length()`, `head()`, `str()` are a simple way to show the features of
a dataset to validate that data transformation steps are working as expected.
Histograms, boxplots, scatterplots and other simple charts are similarly effective.

@@

Journals are partly responsible, as their editorial policies influence author behaviours [@Han2017-pz].
Except for a few outliers, verification of computational reproducibility is not in the scope of peer
review, but automating this in the publication process should be considered.
As others have noted, it is in the interests of funders to provide the infrastructure to facilitate this
[@Peng2011-ia].

@@

Individual researchers also face difficulties.
Life science graduates often receive inadequate training in data science that places them in a
disadvantage considering the data intensive nature of contemporary biology.
This is compounded by the strong demand for data science experts in industry and government which
leads to a deficit of academics fluent in data science
[@Committee_on_Envisioning_the_Data_Science_Discipline_The_Undergraduate_Perspective2018-pe].
This makes it challenging for organisations to recruit and retain capable data analysts and cultivate
a community of practice.
Addressing this problem is not straightforward.
It requires ehancement of reproducible data science skills in undergraduate life science courses,
investment in upskilling the current workforce and offering working conditions that are competitive
with other sectors.
Specific workshop-based training for current researchers could help [@Williams2023-ic].
The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
[@Wilson2014-mh].

A lack of documentated protocols and tutorial materials for highly reproducible analysis makes it
difficult for researchers to confidently update their code development practices.
Expert bioinformaticians can help by publishing comprehensive beginner-friendly step-by-step
guides for frequently used routines.
This approach is highly scalable and the reach of open-access publishing ensures these materials are
broadly accesible.
For example, our group has just released such a protocol that walks novice users through the process
of implementing functional enrichment analysis of gene expression data with the five pillars principles
@Ziemann2023-nm.
More guides like this are needed to make extreme reproducibility mainstream.
A compendium of publicly available learning materials around the five pillar concepts are
provided in the supplementary information.

@@

As the chemotherapeutics case study highlights, there are significant risks to poorly conducted
bioinformatics research; and current standards are falling short.
If adopted widely, the transparency and reproducibilty measures proposed here will reduce the chance of
such disasters from happening again.
Researchers, their institutions, publishers and funders each have a major role to play in fighting the
reproducibility crisis by encouraging highly reproducible research practices.

@@

## Acknowledgements

>>>>

* Correspondence: m.ziemann@deakin.edu.au

| `Author      ` |  `ORCID             ` |
| --- | --- |
| Mark Ziemann | 0000-0002-7688-6974 |
| Pierre Poulain | 0000-0003-4177-3619 |
| Anusuiya Bora | 0009-0006-2908-1352 |

@@

In research, computational reproducibility is the ability to use the materials from a past study
(such as data, code and documentation) to regenerate the outputs including figures and tables to
confirm the study's findings @Goodman2016-wi.
Reproducibility is only the first step towards overall reliability, for example a study may be
reproducible, but suffer from analytical problems that invalidate the results.
If a study is reproducible, then at least these analytical issues can be more easily identified and
rectified.
Therefore, reproducibility is a necessary stepping-stone in the framework of cumulative science.
Reproducibility should enhance the overall reliability of computational research
including replicability and robustness.
Replicability being defined as the ability to repeat the entire study and come up with similar overall
results.
Robustness refers to the ability for the findings to be generalizable to other situations
@Nosek2022-zc.

@@

For example, in 2006 an article entitled "Genomic signatures to guide the use of chemotherapeutics"
generated a great deal of interest as it was an early application of high throughput transcriptomics
in the prediction of individual patient responses to different chemotherapies @Potti2006-yk.
After observing some unusual features of the patient group, Baggerly and Coombes attempted
reproduction of some of the key findings @Baggerly2010-nr.
Without the help of scripted workflows to guide the re-analysis, the team used forensic bioinformatic
techniques to piece together how the dataset was originally analyzed.
Their investigations found a litany of issues.
Firstly, the labeling of patients in the test set as "responders" or "non-responders" had been reversed
in the original analysis.
Secondly, some of the patients were included more than once (some up to four times) in the analysis,
likely to cause major distortions in results.
Confusingly, some of the reused data has inconsistent grouping; ie: some of the samples are labeled
both sensitive and resistant.
Additional errors include two cases where results (charts) were ascribed to the wrong drug.
Baggerly and Coombes highlight that such mistakes can inadvertently occur when conducting unscripted
data analysis such as using spreadsheets, and these problems can be obscured by a lack of documentation.
The article underwent two corrigenda, but was ultimately retracted in 2011, as the authors were not
able to reproduce the findings themselves due to "corruption of several validation data sets"
@Potti2011-ut.
As the array findings were the basis for clinical trials where patients were allocated to treatments,
the flawed data analysis may have harmed patients given the wrong drug in the
period 2007-2010.
In 2010, Duke University terminated the trials and suspended the lead author, Dr Anil Potti, who
later resigned.
Duke was served eight lawsuits by families of affected patients seeking compensation for exposure to
harmful and unnecessary chemotherapy, which were settled out of court @Dyer2015-my.
This worst-case scenario emphasizes that computational reproducibility is crucial for
translating bioinformatics research into real-world outcomes.

@@

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerized applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerization system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
[@Kurtzer2017-zq], Docker remains the most widely used containerization system (according to GitHub
stars as of August 2023).
Docker has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei; @Mangul2019-xp; @Krafczyk2021-xv].

@@

Literate programming also provides an opportunity to record sanity checks in an output report.
In R, commands like `dim()`, `length()`, `head()`, `str()` are a simple way to show the features of
a dataset to validate that data transformation steps are working as expected.
Histograms, box plots, scatter plots and other simple charts are similarly effective.

@@

Journals are partly responsible, as their editorial policies influence author behaviors [@Han2017-pz].
Except for a few outliers, verification of computational reproducibility is not in the scope of peer
review, but automating this in the publication process should be considered.
As others have noted, it is in the interests of funders to provide the infrastructure to facilitate this
[@Peng2011-ia].

@@

Individual researchers also face difficulties.
Life science graduates often receive inadequate training in data science that places them in a
disadvantage considering the data intensive nature of contemporary biology.
This is compounded by the strong demand for data science experts in industry and government which
leads to a deficit of academics fluent in data science
[@Committee_on_Envisioning_the_Data_Science_Discipline_The_Undergraduate_Perspective2018-pe].
This makes it challenging for organizations to recruit and retain capable data analysts and cultivate
a community of practice.
Addressing this problem is not straightforward.
It requires enhancement of reproducible data science skills in undergraduate life science courses,
investment in up-skilling the current workforce and offering working conditions that are competitive
with other sectors.
Specific workshop-based training for current researchers could help [@Williams2023-ic].
The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
[@Wilson2014-mh].

A lack of documented protocols and tutorial materials for highly reproducible analysis makes it
difficult for researchers to confidently update their code development practices.
Expert bioinformaticians can help by publishing comprehensive beginner-friendly step-by-step
guides for frequently used routines.
This approach is highly scalable and the reach of open-access publishing ensures these materials are
broadly accessible.
For example, our group has just released such a protocol that walks novice users through the process
of implementing functional enrichment analysis of gene expression data with the five pillars principles
@Ziemann2023-nm.
More guides like this are needed to make extreme reproducibility mainstream.
A compendium of publicly available learning materials around the five pillar concepts are
provided in the supplementary information.

@@

As the chemotherapeutics case study highlights, there are significant risks to poorly conducted
bioinformatics research; and current standards are falling short.
If adopted widely, the transparency and reproducibility measures proposed here will reduce the chance of
such disasters from happening again.
Researchers, their institutions, publishers and funders each have a major role to play in fighting the
reproducibility crisis by encouraging highly reproducible research practices.

@@

## Acknowledgments
